A simplified dynamic bioenergetic model for coral-*Symbiodinium* symbioses

### 2016-10-06 JTB_ms/JTB_ms.Rmd
updates from UCSB

bibliography: library.bib
output: 
  rticles::elsevier_article:
    fig_caption: yes
---
```{r setup, include=FALSE}
#knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load, include=FALSE}
# Source default parameters and functions
sapply(list.files(path="../../R", pattern="*.R", full.names=T), source, .GlobalEnv)
```

@@

The nutritional exchange between corals and *Symbiodinium* directly underlies the capacity of corals to build coral reef ecosystems, worth trillions of US Dollars annually [@Costanza:2014ex]. However, the complex symbiotic metabolism of corals is vulnerable to disruption by numerous anthropogenic environmental perturbations, jeopardizing their future persistence. In order to understand and predict coral responses to complex changes in the environment, a mechanistic understanding of how multiple interacting factors drive the individual and emergent physiology of both symbiotic partners is necessary. Such a task is well suited for theoretical modeling frameworks such as Dynamic Energy Budget (DEB) theory [@Kooijman:2010vd], although the complexity of such theory makes these efforts inaccessible to many biologists [@Jager:2013bj]. In order to bridge this gap, we present here a simplified dynamic bioenergetic model for coral-*Symbiodinium* symbioses that aims to mechanistically integrate the impacts of complex environmental change on the physiological ecology of reef corals.

In reef coral symbioses, intracellular *Symbiodinium* translocate photosynthetically-fixed carbon to support coral metabolism, and utilize the animal’s metabolic waste products, including nitrogenous compounds and carbon dioxide, in return [@Muscatine:1977p4220]. Previous application of DEB theory to this syntrophic system [@Muller:2009io] demonstrated a stable symbiotic relationship and qualitatively realistic growth and biomass ratios across gradients of ambient irradiance, nutrients, and food. This model assumed that 1) *Symbiodinium* has priority access to carbon through photosynthesis, 2) the coral animal has priority access to dissolved nitrogen through contact with seawater, and 3) each partner shares with the other only what it cannot use itself. In its simplest form, this principle of sharing the surplus sufficiently describes diverse syntrophic interactions among organs and organisms (e.g., trees, duckweeds, corals), suggesting the mechanism is mathematically and evolutionarily robust (Nisbet et al., submitted).

While the work of @Muller:2009io applying DEB formalism to this system represents the most significant theoretical contribution in coral symbiosis research to date, we aim to strengthen the role of theory and broaden its potential application to corals in three primary ways:

@@

2. _Reduce theoretical and mathematical complexity._ Following the logic of @Jager:2013bj, we exclude certain features of formal DEB models in order to capture behaviors of interest with the simplest possible formulation. Here, we present a model without reserves, maturity, or reproduction [see @Kooijman:2010vd]. This formulation restricts the model's scope to the bioenergetics of growth and symbiosis dynamics in adult corals, but greatly reduces theoretical complexity and parameter numbers, which is advantageous given the relative paucity of data for corals. However, our primary motivation for reducing complexity was to increase accessibility and applicability for biologists and ecologists without requiring significant expertise in DEB theory.

3. _Provide well-documented, open-access code._ In order to facilitate the continued development and application of theoretical modeling tools for coral symbioses, we provide open access to the model in the form of detailed, commented code written in the R language [@RALanguageandEn:2014wf]. With an accessible and modular framework, we envision this as a resource for futher development by the scientific community to include additional complexity and problem-specific components. The R language was chosen because it is freely available and in common use by biologists and ecologists, to widen the audience for this work.

@@

# Model description
In this model of coral-algal symbiosis, carbon and nitrogen are acquired by each partner and used to construct biomass. A graphical representation of the model is presented in Fig. 1. We use C-moles as the unit of biomass for consistency with the rigorous mass balance of DEB theory: 1 C-mole is equivalent to the amount of biomass containing 1 mole of Carbon atoms. Host biomass ($H$), symbiont biomass ($S$), and prey biomass ($X$) have fixed, but different, molar N:C ratios (Table 1). Carbon and nitrogen are combined to produce biomass by synthesizing units (SU), which are mathematical specifications of the formation of a product from two substrates; we use the "parallel complementary" formulation of @Kooijman:2010vd to specify these fluxes. The two state variables in this dynamical system are symbiont biomass and coral biomass; because resources are acquired proportionally to surface area (and surface area is assumed proportional to volume for corals (i.e., they are "V1-morphs" in DEB terminology [@Kooijman:2010vd])), biomass increases exponentially during growth. The rate of increase in coral biomass (i.e. growth) and the ratio of symbiont to host biomass are the (i.e. symbiosis dynamics) are the responses of interest of the system. Below we describe the formulation of each flux involved in producing these responses.

Table 1

| Symbol  | Description                            | Value  | Units              |
| ------- | -------------------------------------- | ------ | ------------------ |
| $n_{NH}$ | N:C molar ratio in host biomass | 0.19 | -- |
| $n_{NS}$ | N:C molar ratio in symbiont biomass | 0.2 | -- |
| $n_{NX}$ | N:C molar ratio in prey biomass | 0.13 | -- |
| $j_{HT}^0$ | Specific turnover rate of host biomass | 0.03 | $d^{-1}$ |
| $j_{ST}^0$ | Specific turnover rate of symbiont biomass | 0.03 | $d^{-1}$ |
| $\sigma_{NH}$ | Proportion host nitrogen turnover recycled | 0.9 | -- |
| $\sigma_{CH}$ | Proportion host carbon turnover recycled | 0.9 | -- |
| $\sigma_{NS}$ | Proportion symbiont nitrogen turnover recycled | 0.9 | -- |
| $\sigma_{CS}$ | Proportion symbiont carbon turnover recycled | 0.9 | -- |
| $j_{Xm}$ | Maximum specific feeding rate of host | 0.1292 | $molX \cdot CmolH^{-1} \cdot d^{-1}$ |
| $K_X$ | Half-saturation constant for prey uptake by host | 20e-6 | $molX \cdot L^{-1}$ | 
| $j_{Nm}$ | Maximum specific DIN uptake rate by host | 0.048 | $molN \cdot CmolH^{-1} \cdot d^{-1}$ |
| $K_N$ | Half-saturation constant for DIN uptake by host | 0.46e-6 | $molN \cdot L^{-1}$ | 
| $j_{CO_2}^p$ | Passive CO~2~ delivery to symbiont | 4.04e-3 | $molC \cdot CmolH^{-1} \cdot d^{-1}$ |
| $j_{CO_2}^a$ | Active CO~2~ delivery to symbiont | 0.32--18.0 | $molC \cdot CmolH^{-1} \cdot d^{-1}$ |
| $j_{HGm}$ | Maximum specific growth rate of host | 1 | $d^{-1}$ |
| $n_{LC}$ | Quantum yield of photosynthesis | 0.1 | $molC \cdot mol ph^{-1}$ |
| $\bar{a}^*$ | Effective light-absorbing cross-section of symbiont | 1.34 | $m^2 \cdot CmolS^{-1}$ |
| $j_{NPQ}$ | Non-photochemical quenching capacity of symbiont | 40 | $mol ph \cdot CmolS^{-1} \cdot d^{-1}$|
| $k_{ROS}$ | Excess photon energy that doubles ROS prod. | 40-80 | $mol ph \cdot CmolS^{-1} \cdot d^{-1}$|
| $k$ | Exponent on ROS production rate | 1 | -- |
| $j_{CPm}$ | Maximum specific photosynthesis rate of symbiont | 2.8 | $molC \cdot CmolS^{-1} \cdot d^{-1}$ |
| $j_{SGm}$ | Maximum specific growth rate of symbiont | 0.25 | $d^{-1}$ |

Also include tables of fluxes and environmental inputs?

![Graphical representation of coral-algal symbiosis model.](../img/model.png)

@@

The coral animal acquires both carbon and nitrogen from feeding on prey from the environment. Prey acquisition is specified by Michaelis-Menten kinetics using a maximum area-specific feeding rate and half-saturation constant:

@@

Additionally, the coral animal can acquire nitrogen from the surrounding seawater. This nitrogen source is assumed to represent ammonium, the primary form utilized by corals [@Wang:1998p128; @Yellowlees:2008p331]. The uptake of nitrogen from the environment is specified by Michaelis-Menten kinetics using a maximum area-specific uptake rate and half-saturation constant:

@@

Coral biomass formation is specified by a parallel complementary SU that combines carbon and nitrogen to form biomass. In addition to the carbon and nitrogen acquired by direct uptake (Eqs. 1 and 2), the coral recycles a portion of the nitrogen liberated by biomass turnover ($r_{NH}=\sigma_{NH}j_{HT}^0$), and receives surplus fixed carbon shared by the symbiont ($\rho_C$), such that the total biomass formation is specified as:

@@

The amount of nitrogen input to the coral biomass SU in excess of what is actually consumed in biomass formation (i.e., surplus nitrogen, or the "rejection flux" in SU terminology) is then made available to the symbiont, and is specified as:

@@

Due to the inherent inefficiency of the parallel complementary SU formulation, there will always be some nitrogen shared with the symbiont even when coral biomass formation is strongly nitrogen-limited. Likewise, there is always a non-zero rejection flux of carbon from the coral biomass SU, which is assumed to be lost to the environment.

@@

The symbiont produces fixed carbon through photosynthesis, a process represented here by a single SU with two substrates: light (photons) and inorganic carbon (CO~2~). The amount of light absorbed by the symbiont depends on the scalar irradiance at the site of light absorption, which is modified substantially relative to external downwelling irradiance owing to multiple scattering by the coral skeleton and self-shading by surrounding symbionts [@Enriquez:2005p142; @Marcelino:2013hz]. We used data from @Marcelino:2013hz to empirically derive the ratio of internal scalar irradiance to external downwelling irradiance as a function of symbiont density (expressed as symbiont to host biomass ratio), and subsequently multiply this quantity by the external downwelling irradiance $L$ and the effective light-absorbing surface area of symbiont biomass $ds$ to specify the amount of light absorbed:

\begin{equation} j_L = \big[1.26 + 1.39 \cdot \exp(-6.48 \cdot {S \over H})\big] \cdot L \cdot \bar{a}^* \end{equation}

We then specify two pathways for input of inorganic carbon to the photosynthesis SU: 1) passive diffusion of CO~2~ from the external environment, and 2) active delivery of CO~2~ to the symbiont by the host. The passive flux ensures that some CO~2~ is always available to photosynthesis, and the active flux encompasses the potentially diverse mechanisms by which the host may enhance CO~2~ availability for the symbiont, including active transport of bicarbonate, carbonic anhydrase-catalyzed conversion of bicarbonate to CO~2~ to promote diffusion toward the symbiont, and acidification of the symbiosome to increase localized CO~2~ concentrations. Since the host physically separates the symbiont from the external environment, both the passive and active flux rates are proportional to host surface area. A more rigorous, mechanistic model of inorganic carbon processing would require spatially explicit internal pools accounting for pH and carbon speciation, which is beyond the current scope of this work. Instead, the specification of CO~2~ delivery rates offers the user the opportunity to compare different rates of CO~2~ delivery that may characterize different coral species [@Wooldridge:2014di]. The input of CO~2~ to the photosynthesis SU is therefore specified as:

@@

In addition to the inputs CO~2~ specified in Eqs. 7, additional CO~2~ representing the metabolic production of CO~2~ from host and symbiont biomass turnover ($r_{CH}=\sigma_{CH}j_{HT}^0; r_{CS}=\sigma_{CS}j_{ST}^0$) is made available to the photosynthesis SU. fixed carbon is produced by the photosynthesis SU according to:

@@

where $j_{CPm}$ is the maximum specific rate of photosynthesis, and $c_{ROS}$ is the photooxidative stress multiplier (see below). Dividing by $c_{ROS}$ causes the rate of photosynthesis to decline in response to photo-oxidative stress, a phenomenon known as photoinhibition.

@@

This excess light energy must be quenched by alternative pathways in order to prevent photo-oxidative damage. *Symbiodinium* may utilize a variety of pathways for non-photochemical quenching [NPQ; @Roth:2014wf], which we collect in a total capacity for NPQ as a parameter of the symbiont ($j_{NPQ}$). If light energy exceeds the total capacity of both carbon fixation and NPQ, then damaging reactive oxygen species (ROS) may be produced. We represent this as a scaled flux of ROS $c_{ROS}$, which takes a value of 1 when all light absorbed is quenched by photochemistry and NPQ, and increases as the amount of excess excitation energy increases.

\begin{equation} c_{ROS} = 1 + \bigg[\bigg({{j_{eL} - j_{NPQ}} \over k_{ROS}}\bigg)^k\bigg]_+ \end{equation}

where $j_{NPQ}$, $k_{ROS}$, and $k$ are parameters of the symbiont that determine the onset and rate of ROS production. Importantly, $c_{ROS}$ as specified here is not a function of absolute light absorption, but rather the amount of excess light energy $j_{eL}$ after accounting for carbon fixation and NPQ. A direct consequence of this formulation is that carbon-limitation of photosynthesis can lead to photo-oxidative stress, a mechanism of biological importance [@Wooldridge:2009p7807] that was not captured by previous representations of photo-oxidative stress [@Eynaud:2011tv]. Moreover, this formulation allows functional diversity among symbiont types to be explored by changing the parameters $j_{NPQ}$, $k_{ROS}$, and $k$.

Carbon fixed by photosynthesis ($j_{CP}$) is then used in conjunction with nitrogen shared by the host ($\rho_N$) and a proportion of nitrogen recycled from symbiont biomass turnover ($r_{NS}=\sigma_{NS}j_{ST}^0$) to build new symbiont biomass, following the SU equation:

@@

The rejection flux of carbon from this SU represents the amount of fixed carbon produced by photosynthesis in excess of what can be used to produce symbiont biomass; this surplus $\rho_C$ is translocated to the coral host:

@@

\begin{equation} j_{ST} = j_{ST}^0(1 + 5 \cdot (c_{ROS}-1)) \end{equation}

This second component of symbiont biomass loss can represent both photodamage and/or symbiont expulsion (i.e., bleaching), both of which happen in response to higher levels or ROS production. The constant 5 is included to increase biomass loss in response to ROS. (Note that recycling of symbiont biomass turnover ($rNS$ and $rCS$) only occurs based on the basal maintenance related turnover (i.e., $j_{ST}^0$), and not the bleaching-related biomass loss, as this loss represents biomass being damaged or being expelled from the holobiont).

## Model state equations
Finally, the balance equations representing the specific growth rates of symbiont and host biomass over time can be expressed as:

@@

A time-stepping Euler method was used to solve the state equations since the production and rejection fluxes of the SUs are imiplicitly defined. Specifically, the rejection fluxes of carbon and nitrogen from the symbiont and host biomass SUs act as reciprocal input fluxes to the other SU. In addition, the rejection flux of excitation energy from the photosynthesis SU acts to reduce its own production flux (i.e., photoinhibition), and hence a discretized time-stepping procedure was necessary. A vector of time values was created for each simulation run, along which dynamic environmental forcing functions (irradiance, DIN, and prey abundance) can be designed. These vectors, along with initial values of symbiont and host biomass, then serve as input to the time-stepping function, which solves for the current system state using values of the previous system state where necessary. A default time step of 0.1 days was used for all simulations, which were performed using R code that is available in the data repository accompanying this article: [github.com/jrcunning/Rcoral](github.com/jrcunning/Rcoral).

# Model behavior evaluation

To analyze qualitative behavior, we ran the model to steady state across gradients of external irradiance and DIN, and plotted the specific growth rate (Fig. 2A) and symbiont to host biomass ratio (Fig. 2B). Results are consistent with observations in corals: specific growth rates are low at very low levels of irradiance and DIN, as these scenarios limit the availability of carbon and nitrogen, respectively, to build biomass. As these resources increase, specific growth increases, reaching its highest levels above 2 µM DIN and ~15 mol photons m^-2^ d^-1^. Peak specific growth rates are around ~0.1 d^-1^, which is consistent with values reported in the literature (@Muscatine:1981jy).

![Steady state values of (A) specific growth (Cmol Cmol^-1^ d^-1^) and (B) the symbiont to host biomass ratio (CmolS CmolH^-1^) across gradients of external irradiance and dissolved inorganic nitrogen. Simulations for each combination of light and nutrients (21 points along each axis) were run for 100 days with a time step of 1 day. Negative steady state growth rates, and corresponding S:H ratios, were set to zero.](../img/Fig2.png)

# Discussion  -  Potential applications/utility

>>>>

bibliography: library.bib
output: 
  rticles::elsevier_article:
    fig_caption: yes
---
```{r setup, include=FALSE}
library(reshape2); library(dplyr)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load, include=FALSE}
# Source metrics reported in text
source("R/reported_metrics.R")
```

@@

The nutritional exchange between corals and *Symbiodinium* directly underlies the capacity of corals to build coral reef ecosystems, worth trillions of US Dollars annually [@Costanza:2014ex]. However, the complex symbiotic metabolism of corals is vulnerable to disruption by numerous anthropogenic environmental perturbations, jeopardizing their future persistence. In order to understand and predict coral responses to complex changes in their environment, a mechanistic understanding of how multiple interacting factors drive the individual and emergent physiology of both symbiotic partners is necessary. Such a task is well suited for theoretical modeling frameworks such as Dynamic Energy Budget (DEB) theory [@Kooijman:2010vd], although the complexity of such theory makes these efforts inaccessible to many biologists [@Jager:2013bj]. In order to bridge this gap, we present here a simplified dynamic bioenergetic model for coral-*Symbiodinium* symbioses that aims to mechanistically integrate the impacts of complex environmental change on the physiological performance of reef corals, including responses to environmental stress.

In reef coral symbioses, intracellular *Symbiodinium* translocate photosynthetically fixed carbon to support coral metabolism, while the animal host provides access to inorganic nutrients and carbon dioxide [@Muscatine:1977p4220]. Previous application of DEB theory to this syntrophic system [@Muller:2009io] demonstrated a stable symbiotic relationship and qualitatively realistic growth and biomass ratios across gradients of ambient irradiance, nutrients, and food. This model assumed that 1) *Symbiodinium* has priority access to fixed carbon through photosynthesis, 2) the coral animal has priority access to dissolved nitrogen through contact with seawater, and 3) each partner shares with the other only what it cannot use for its own growth. In its simplest form, this principle of sharing the surplus is sufficient to describe the dynamics of diverse syntrophic organs and organisms (e.g., trees, duckweeds, corals), suggesting the mechanism is mathematically and evolutionarily robust (Nisbet et al., in prep.).

While the formal DEB model of @Muller:2009io represents the most significant theoretical contribution in coral symbiosis research to date, we aim to strengthen the role of theory and broaden its potential application in three primary ways:

@@

2. _Reduce theoretical and mathematical complexity._ Following the logic of @Jager:2013bj, we exclude certain features of formal DEB models in order to capture behaviors of interest with the simplest possible formulation. Here, we present a model without reserves, maturity, or reproduction [see @Kooijman:2010vd]. This formulation restricts the model's scope to the bioenergetics of growth and symbiosis dynamics in adult corals (i.e., reproduction, larval stages, and metamorphosis are not considered), but greatly reduces theoretical complexity and parameter numbers, which is advantageous given the relative paucity of data for corals. However, our primary motivation for reducing complexity was to increase accessibility and applicability for biologists and ecologists without requiring significant expertise in DEB theory.

3. _Provide well-documented, open-access code._ In order to facilitate the continued development and application of theoretical modeling tools for coral symbioses, we provide open access to the model in the form of detailed, commented code written in the R language [@RALanguageandEn:2014wf]. With an accessible and modular framework, we envision this as a resource for futher development by the scientific community to include additional complexity and problem-specific components. To widen the audience for this work, we chose the R language because it is freely available and in common use by biologists and ecologists.

@@

# Model description
In this dynamic bioenergetic model of coral-algal symbiosis, carbon and nitrogen are acquired by each partner and used to construct biomass. A graphical representation of the model is presented in Fig. 1, and each model flux and parameter are defined in Tables 1 and 2, respectively. We use C-moles as the unit of biomass for consistency with the rigorous mass balance of DEB theory: 1 C-mole is equivalent to the amount of biomass containing 1 mole of Carbon atoms. Host biomass ($H$), symbiont biomass ($S$), and prey biomass ($X$) have fixed, but different, molar N:C ratios (Table 2). Carbon and nitrogen are combined to produce biomass by synthesizing units (SU), which are mathematical specifications of the formation of a product from two substrates; we use the parallel complementary formulation of @Kooijman:2010vd to specify these fluxes. The two state variables are symbiont biomass and coral biomass; because resources are acquired proportionally to surface area, and surface area is proportional to volume (i.e., corals are "V1-morphs" in DEB terminology [@Kooijman:2010vd]), biomass increases exponentially during growth (indeed, corals grow exponentially [@Bak:1976bv]). The specific growth rate and the ratio of symbiont to host biomass are the responses of interest of the system. Below we describe the formulation of each flux involved in producing these responses.

Table 1. Model fluxes

| Symbol    | Description                            | Units                | Eq. no. |
| --------- | -------------------------------------- | -------------------- | ------- |
| $j_X$     | Prey uptake rate                       | molX CmolH^-1^ d^-1^ | 3 |
| $j_N$     | Nitrogen uptake rate                   | molN CmolH^-1^ d^-1^ | 4 |
| $j_{HG}$  | Host biomass formation rate            | CmolH CmolH^-1^ d^-1^| 5 |
| $r_{NH}$  | Recycled nitrogen from host turnover   | molN CmolH^-1^ d^-1^ | 6 |
| $\rho_N$  | Nitrogen shared with the symbiont      | molN CmolH^-1^ d^-1^ | 7 |
| $j_L$     | Light absorption rate                  | molph CmolS^-1^ d^-1^| 9 |
| $j_{CO_2}$| CO~2~ input to photosynthesis          | molCO~2~ CmolH^-1^ d^-1^| 10 |
| $r_{CH}$  | Recycled CO~2~ from host turnover      | molCO~2~ CmolH^-1^ d^-1^| 11 |
| $r_{CS}$  | Recycled CO~2~ from symbiont turnover  | molCO~2~ CmolS^-1^ d^-1^| 12 |
| $j_{CP}$  | Photosynthesis rate                    | molC CmolS^-1^ d^-1^ | 13 |
| $j_{eL}$  | Light energy in excess of photochemistry| molph CmolS^-1^ d^-1^| 14 |
| $j_{NPQ}$ | Total capacity of NPQ                  | molph CmolS^-1^ d^-1^ | 15 |
| $c_{ROS}$ | ROS production proportional to baseline | -- | 16 |
| $r_{NS}$  | Recycled nitrogen from symbiont turnover| molN CmolS^-1^ d^-1^| 17 |
| $j_{SG}$  | Symbiont biomass formation rate        | CmolS CmolS^-1^ d^-1^| 18 |
| $\rho_C$  | Fixed carbon shared with host          | molC CmolS^-1^ d^-1^ | 19 |
| $j_{ST}$  | Symbiont turnover rate                 | CmolS CmolS^-1^ d^-1^| 20 |

Table 2. Model parameters

| Symbol  | Description                            | Value  | Units              |
| ------- | -------------------------------------- | ------ | ------------------ |
| $n_{NH}$ | N:C molar ratio in host biomass | 0.19 | -- |
| $n_{NS}$ | N:C molar ratio in symbiont biomass | 0.2 | -- |
| $n_{NX}$ | N:C molar ratio in prey biomass | 0.13 | -- |
| $j_{HT}^0$ | Specific turnover rate of host biomass | 0.03 | CmolH CmolH^-1^ d^-1^ |
| $j_{ST}^0$ | Specific turnover rate of symbiont biomass | 0.03 | CmolS CmolS^-1^ d^-1^ |
| $\sigma_{NH}$ | Proportion host nitrogen turnover recycled | 0.9 | -- |
| $\sigma_{CH}$ | Proportion host carbon turnover recycled | 0.9 | -- |
| $\sigma_{NS}$ | Proportion symbiont nitrogen turnover recycled | 0.9 | -- |
| $\sigma_{CS}$ | Proportion symbiont carbon turnover recycled | 0.9 | -- |
| $j_{Xm}$ | Maximum specific feeding rate of host | 0.1292 | molX CmolH^-1^ d^-1^ |
| $K_X$ | Half-saturation constant for prey uptake by host | 20e-6 | molX L^-1^ | 
| $j_{Nm}$ | Maximum specific DIN uptake rate by host | 0.048 | molN CmolH^-1^ d^-1^ |
| $K_N$ | Half-saturation constant for DIN uptake by host | 0.46e-6 | molN L^-1^ | 
| $j_{CO_2}^p$ | Passive CO~2~ delivery to symbiont | 4.04e-3 | molC CmolH^-1^ d^-1^ |
| $j_{CO_2}^a$ | Active CO~2~ delivery to symbiont | 0.32--18.0 | molC CmolH^-1^ d^-1^ |
| $j_{HGm}$ | Maximum specific growth rate of host | 1 | CmolH CmolH^-1^ d^-1^ |
| $n_{LC}$ | Quantum yield of photosynthesis | 0.1 | molC mol ph^-1^ |
| $\bar{a}^*$ | Effective light-absorbing cross-section of symbiont | 1.34 | m^2^ CmolS^-1^ |
| $k_{NPQ}$ | NPQ capacity of symbiont, relative to C-fixation | 4 | -- |
| $k_{ROS}$ | Excess photon energy that doubles ROS production, relative to baseline levels | 40-80 | mol ph CmolS^-1^ d^-1^ |
| $k$ | Exponent on ROS production rate | 1 | -- |
| $j_{CPm}$ | Maximum specific photosynthesis rate of symbiont | 2.8 | molC CmolS^-1^ d^-1^ |
| $j_{SGm}$ | Maximum specific growth rate of symbiont | 0.25 | CmolS CmolS^-1^ d^-1^ |
| $b$ | Scaling parameter for strength of bleaching response | 3 | -- |

![Graphical representation of coral-algal symbiosis model. Light, CO~2~, prey, and DIN are acquired from the external environment proportional to the biomass of the partner indicated by the black box for uptake. Mass fluxes (see Table 1 for definitions) are represented by $j$'s with subscripts indicating the type of mass, and in some cases the process (e.g., $j_{CP}$ is the flux of carbon produced by photosynthesis), and $\rho$'s indicate fluxes that are shared by one partner with the other. Parallel complementary synthesizing units (SUs) are represented by large circles, and rejection fluxes from these SUs are indicated by dashed lines. $c_{ROS}$ is a proportional rate that impacts other model fluxes by inhibition or acceleration. Recycled mass fluxes from biomass turnover are not shown for clarity (but see Table 1 for definitions).](../img/Fig1.png)

## State equations
The balance equations for symbiont and host biomass are expressed as:

These are expressed per unit of symbiont and host biomass, respectively. The specific biomass growth and turnover rates that define these balance equations are produced by combinations of the individual model fluxes (see Table 1 for definitions and units), which are each expressed as mass-specific rates (e.g., per C-mole of symbiont or host biomass). When necessary, conversions between symbiont-mass-specific and host-mass-specific rates are accomplished by multiplying or dividing by the symbiont:host biomass ratio.

@@

The coral animal acquires both carbon and nitrogen from feeding on prey from the environment. Prey acquisition is specified by Michaelis-Menten kinetics (i.e., a Holling type II function) with a maximum feeding rate $j_{Xm}$ and half-saturation constant $K_X$:

@@

Additionally, the coral animal acquires nitrogen dissolved in the surrounding seawater, which is assumed to represent ammonium, the primary form utilized by corals [@Yellowlees:2008p331]. This gives the host (rather than the symbiont) priority in nitrogen utilization; this capacity is supported by experimental evidence [@Wang:1998p128] and is consistent with the physical arrangment of the partners, where the host is in direct contact with the external environment. The uptake of nitrogen from the environment is thus specified by Michaelis-Menten kinetics using a maximum uptake rate $j_{Nm}$ and half-saturation constant $K_N$:

@@

Coral biomass formation is then specified by a parallel complementary SU that combines carbon and nitrogen, according to:

@@

where $\rho_C$ is the surplus fixed carbon shared by the symbiont (see below), and $r_{NH}$ is the recycled portion of the nitrogen liberated by host biomass turnover:

\begin{equation} r_{NH}=\sigma_{NH}j_{HT}^0 \end{equation}

The amount of nitrogen input to the coral biomass SU in excess of what is actually consumed in biomass formation (i.e., surplus nitrogen, or the rejection flux[^1] of the SU) is then made available to the symbiont:

@@

[^1]: Rejection fluxes must always be positive, and hence are specified with the notation $(x)_+$, which means $max(x, 0)$.

Due to the inherent inefficiency of the parallel complementary SU formulation, there is always some nitrogen shared with the symbiont even when coral biomass formation is strongly nitrogen-limited. Likewise, there is always a non-zero rejection flux of carbon from the coral biomass SU, which is assumed to be lost to the environment.

@@

The symbiont produces fixed carbon through photosynthesis, a process represented here by a single SU with two substrates: light (photons) and inorganic carbon (CO~2~). The amount of light absorbed by the symbiont depends on the scalar irradiance at the site of light absorption, which is modified substantially relative to external downwelling irradiance owing to multiple scattering by the coral skeleton and self-shading by surrounding symbionts [@Enriquez:2005p142; @Marcelino:2013hz]. We used data from @Marcelino:2013hz to empirically derive an "amplification factor" $A$ (the ratio of internal scalar irradiance to external downwelling irradiance) as a function of symbiont density (expressed as symbiont to host biomass ratio), which is specified as:

\begin{equation} A = 1.26 + 1.39 \cdot \exp(-6.48 \cdot {S \over H}) \end{equation}

This amplification factor is then multiplied by the external downwelling irradiance $L$ and the effective light-absorbing surface area of symbiont biomass $\bar{a}^*$ to specify the total amount of light absorbed:

\begin{equation} j_L =  A \cdot L \cdot \bar{a}^* \end{equation}

We then specify two pathways for input of inorganic carbon to the photosynthesis SU: 1) passive diffusion of CO~2~ from the external environment ($j_{CO_2}^p$), and 2) active delivery of CO~2~ to the symbiont by the host ($j_{CO_2}^a$). The passive flux ensures that some CO~2~ is always available to photosynthesis, and the active flux encompasses potentially diverse mechanisms by which the host may enhance CO~2~ availability for the symbiont, including active transport of bicarbonate, carbonic anhydrase-catalyzed conversion of bicarbonate to CO~2~ to promote diffusion toward the symbiont [@Tansik:2015cm], and acidification of the symbiosome to increase localized CO~2~ concentrations around the symbiont [@Barott:2014gx]. Since the host physically separates the symbiont from the external environment, both the passive and active flux rates are proportional to host biomass. A more rigorous, mechanistic model of inorganic carbon processing would require spatially explicit internal pools accounting for pH and carbon speciation, which is beyond the current scope of this work. Instead, the specification of fixed passive and active CO~2~ input rates, $j_{CO_2}^p$ and $j_{CO_2}^a$, offers the user the opportunity to compare different rates of CO~2~ delivery that may characterize different coral species [@Wooldridge:2014di]. The input of CO~2~ to the photosynthesis SU is therefore specified as:

@@

Additional metabolic CO~2~ recycled from both the host and symbiont biomass turnover is also made available to the photosynthesis SU, according to:

\begin{equation} r_{CH}=\sigma_{CH}j_{HT}^0 \end{equation}
\begin{equation} r_{CS}=\sigma_{CS}j_{ST}^0 \end{equation}

Fixed carbon is then produced by the photosynthesis SU according to:

@@

where $j_{CPm}$ is the maximum specific rate of photosynthesis, and $c_{ROS}$ is the relative rate of reactive oxygen species production (see below). Dividing the photosynthetic rate by $c_{ROS}$ causes a decline in response to photooxidative stress (i.e., photoinhibition).

@@

This excess light energy must be quenched by alternative pathways in order to prevent photooxidative damage [@Powles:1984tm]. *Symbiodinium* utilize a variety of pathways for non-photochemical quenching [NPQ; @Roth:2014wf], which we collect in a relative NPQ capacity specified as a parameter of the symbiont ($k_{NPQ}$). NPQ capacity may also be susceptible to reduction by photoinhibition, so the total NPQ capacity is specified as:

\begin{equation} j_{NPQ}= k_{NPQ} \cdot c_{ROS}^{-1} \end{equation}

If light energy further exceeds the total capacity of both carbon fixation and NPQ, then reactive oxygen species (ROS) are produced. We represent this as a relative quantity $c_{ROS}$, which takes a value of 1 when all light absorbed can be quenched by photochemistry and NPQ, and increases as the amount of excess excitation energy further increases:

\begin{equation} c_{ROS} = 1 + \bigg({{(j_{eL} - j_{NPQ})_+} \over k_{ROS}}\bigg)^k \end{equation}

where $k_{NPQ}$, $k_{ROS}$, and $k$ are parameters of the symbiont that determine the onset and rate of ROS production. Importantly, $c_{ROS}$ as specified here is not a function of absolute light absorption, but rather the amount of excess light energy, $j_{eL}$, after accounting for carbon fixation and NPQ. A direct consequence of this formulation is that CO~2~-limitation of photosynthesis can lead to ROS production, an important mechanism [@Wooldridge:2009p7807] that was not captured by previous representations of photooxidative stress [@Eynaud:2011tv]. Moreover, this formulation allows functional diversity among symbiont types to be explored by changing the parameters $k_{NPQ}$, $k_{ROS}$, and $k$.

Carbon fixed by photosynthesis ($j_{CP}$) is then combined with nitrogen shared by the host ($\rho_N$) and nitrogen recycled from symbiont biomass turnover

\begin{equation} r_{NS}=\sigma_{NS}j_{ST}^0 \end{equation}

to build new symbiont biomass, following the SU equation:

@@

The rejection flux of carbon from this SU represents the amount of fixed carbon produced by photosynthesis in excess of what can be used to produce symbiont biomass; this surplus, $\rho_C$, is translocated to the coral host:

@@

\begin{equation} j_{ST} = j_{ST}^0(1 + b(c_{ROS}-1)) \end{equation}

This second component of symbiont biomass loss can represent both photodamage and/or symbiont expulsion (i.e., bleaching), both of which occur in response to high levels or ROS production. The parameter $b$ is included to scale biomass loss due to bleaching in response to ROS. (Note that recycling of symbiont biomass turnover ($rNS$ and $rCS$) only occurs based on the maintenance component of turnover (i.e., $j_{ST}^0$), and not the bleaching-related component, as this loss represents biomass that is damaged or expelled from the host).

@@

A time-stepping Euler method was used to solve the state equations since the production and rejection fluxes of the SUs are imiplicitly defined. Specifically, the rejection fluxes of carbon and nitrogen from the symbiont and host biomass SUs act as reciprocal input fluxes to the other SU. In addition, the rejection flux of excitation energy from the photosynthesis SU acts to reduce its own production flux (i.e., photoinhibition), and hence a discretized time-stepping procedure was necessary. A vector of time values was created for each simulation run, along which dynamic environmental forcing functions (irradiance, DIN, and prey abundance) can be designed. These vectors, along with initial values of symbiont and host biomass, then serve as input to the time-stepping function, which solves for the current system state using values of the previous system state where necessary. A time step of 0.1 days was used for all simulations, which were performed using R code that is available in the data repository accompanying this article: [github.com/jrcunning/Rcoral](github.com/jrcunning/Rcoral).

To aid in visualizing model results, we calculated values to indicate the degree to which product formation at an SU was limited by availability of either of its two substrates:

\begin{equation} \log \bigg({{\min(j_{S1}, j_{Pm})} \over {\min(j_{S2}, j_{Pm})}}\bigg) \end{equation}

where $j_{S1}$ and $j_{S2}$ are the specific input fluxes of the two substrates and $j_{Pm}$ is the maximum specific product formation rate, all with units of Cmol Cmol^-1^ d^-1^.

# Sensitivity analysis -- should come after the steady states bc describes changes in steady states

Each default parameter value used in the model (Table 2) is based on carefully selected relevant literature. A supplementary document detailing the derivations of each of these default parameter values is provided. Nevertheless, it is important to evaluate the sensitivity of the model to changes in these parameter values. We measured fractional change in steady state values in response to fractional change in parameter values, relative to their default values (Table 2).
